---
title: "Network analysis"
author: "Wlad Stroukov"
date: "13/10/2021"

output:
  html_document: 
    code_folding: hide
    toc: true
    number_sections: true
    toc_float: true
---

This tutorial was developed by Katherine Ognyanova, www.kateto.=  

# Introduction to network analysis

The full tutorial can be found [here](https://kateto.net/wp-content/uploads/2018/03/R%20for%20Networks%20Workshop%20-%20Ognyanova%20-%202018.pdf)


# Refresher on R data structures
## Vectors
```{r}
v1 <- 1:5  
v2 <- 5:1
v3 <- c(3,1,5,4,2)
cor(v1,v3)


(v1>2) | (v2>0)   # | is the boolean OR, returns a vector.
(v1>2) & (v2>0)   # & is the boolean AND, returns a vector.
(v1>2) || (v2>0)  # || is the boolean OR, returns a single value
(v1>2) && (v2>0)  # && is the boolean AND, ditto

v3
length(v3) <- 10
v3

```

## Factors
```{r}

eye.col.v <- c("brown", "green", "brown", "blue", "blue", "blue")         #vector
eye.col.f <- factor(c("brown", "green", "brown", "blue", "blue", "blue")) #factor
eye.col.v
eye.col.f
levels(eye.col.f)
as.numeric(eye.col.f)
as.numeric(eye.col.v)
as.character(eye.col.f)  
```

## Matrices and Arrays
Matrix 2D vectors
Arrays >2D
```{r}
m <- 1:20 # 20 elements 
m
dim(m) <- c(5,4) # Dimensions set to 5 & 4, so m is now a 5x4 matrix
m

m <-  matrix(data=1:20, nrow=5, ncol=4)
m
m <- matrix(1:20,5,4)
m

m <- cbind(1:5, 5:1, 5:9) # column bind
m
m <- rbind(1:5, 5:1, 5:9) # row bind
m

m <- matrix(1:100,10,10)
m
m[2,3]
m[2,]
m[1:4,2]
m[-1:-3,] # all rows except 1st to 3rd one

m[1,] == m[1,]
m >3

m[m>3]

t(m)
mt <- t(m)
mt

m %*% mt # matrix multiplication
m*mt     # element-wise multiplication


a <- array(data=1:18, dim=c(3,3,2))
a
```
## Lists
```{r}
l1 <-  list(boo=v1,foo=v2,moo=v3,zoo="Animals!") 
l2 <- list(v1,v2,v3,"Animals!")

l3 <- list()
l4 <- NULL

l1["zoo"] # returns a list
l1[["zoo"]] # returns the numeric element in the list (e.g. numeric vector, character, etc)
l1[1]    # 1st element
l1[[1]]
l1$boo   # $ operator is equivalent to [[]]

l3[[1]] <- 11
l3

l4[[3]] <- c(22,23) # coerces empty object into list and adds an element. unassigned elements will be generated and empty (NULL)

l1[[5]] <- "more elements"
l1[[8]] <- 1:5
l1

l1$something <- "A thing"
l1
names(l1) <- c("a","b","c","d","e","f","g","h","i")
l1
```
## DataFrames
```{r}
df1 <- data.frame(ID=1:4,
                  FirstName=c("John", "Jim", "Jane","Jill"),
                  Female=c(F,F,T,T),
                  Age=c(22,33,44,55),
                  stringsAsFactors = T)

df1["FirstName"] # get column
df1$FirstName
df1[["FirstName"]]

df1$FirstName <-  as.vector(df1$FirstName)


df1[1,]
df1[,1]
df1[1:2,3:4]

df1[df1$Age >30 ,2] # names of everyone over 30

mean(df1[df1$Female==T,4]) # Mean age of female entries


```
## Flow control
```{r}
x <- 5; y <- 10
if (x==0) y <- 0 else y <- y/x
y



ASum <- 0; AProd <- 1
for (i in 1:x)
{
  ASum <- ASum +1
  AProd <- AProd * i
}
ASum
AProd
prod(1:x)



x <- 5
while(x > 0){
  print(x)
  x <- x-1
  }



x <-0
repeat{
  print(x) 
  x <- x+1 
  if(x>10) break
}

```
# Networks
```{r}
rm(list=ls())
library(igraph)
```

## Create networks
```{r}
g1 <- graph(edges=c(1,2, 2,3, 3,1), directed=F) # undirected graph with 3 edges. Numbers are interpreted as Vertex IDs: 1->2, 2->3, 3->1
plot(g1)
# class(g1)
# g1
```



```{r}
g2 <- graph(edges=c(1,2, 2,3, 3,1), n=10) # 10 vertices
plot(g2)
# g2
```

```{r}
g3 <- graph(c("John","Jim", "Jim","Jill", "Jill","John", "John","Jill")) # with named vertices the argument "n" appears to be ignored. 
plot(g3)
```
The edge.curved argument of the `plot()` function adds a curve to the edge, but can mask edges due to overlay. There are supposed to be 2 edges from Jim to Jack.
```{r}
g4 <- graph(c("John", "Jim", "Jim", "Jack", "Jim", "Jack", "John", "John"),
            isolates=c("Jesse","Janis","Jennifer", "Justin"))
g4
#plot(g4)
plot(g4, edge.arrow.size=.5, vertex.color="gold", vertex.size=15,
     vertex.frame.color="gray", vertex.label.color="black",
     vertex.label.cex=0.8, vertex.label.dist=2, edge.curved=0.5) 


```
As in the example above, in most R functions, you can use named colors, hex, or rgb values. For example: “blue” “dark red” “#557799” or rgb(.25, .5, .3)) You can see the built-in colors name with `colors()`  

Small graphs can also be generated with a description of this kind: - for undirected tie, +- or -+
for directed ties pointing left & right, ++ for a symmetric tie, and “:” for sets of vertices.
```{r}
plot(graph_from_literal(a----b, b----c)) # the number of dashes doesn't matter
plot(graph_from_literal(a+b, b+c))
plot(graph_from_literal(a+-+b, b+-+c))
plot(graph_from_literal(a+-b, b+-c))


plot(graph_from_literal(a1:a2:a3--b1:b2:b3)) # sets of vertices with undirected edges
plot(graph_from_literal(a1:a2:a3--+b1:b2:b3, c--a1:a2:b3)) # it appears that all edges the either directed or undirected. Otherwise the undirected edges are ignored
plot(graph_from_literal(a1:a2:a3--+b1:b2:b3, c--+a1:a2:b3))

plot(graph_from_literal(a-b-c-d-e-f, a-g-h-b, h-e:f:i, j))
```
## Edge, vertex, and network attributes
```{r}
g4 <- graph(c("John", "Jim", "Jim", "Jack", "Jim", "Jack", "John", "John"),
            isolates=c("Jesse","Janis","Jennifer", "Justin"))
plot(g4)

```
```{r}
E(g4) # Edges
V(g4) # Vertices

g4[] # show as matrix
g4[1,]
```
Add attributes to the network, vertices, or edges:
```{r}
V(g4)$name # automatically generated when the network was created
```
examine attributes
```{r}
edge_attr(g4) # emtpy list
```
Assign new attributes to vertices and edges
```{r}
V(g4)$gender <- c("male","male","male","female","female","female", "male")
E(g4)$type <- "email" # Edge attribute "email" is assigned to all edges
E(g4)$weight <- 10    # Edge weight, setting all existing edges to 10
```

Check Edge attributes
```{r}
edge_attr(g4)
```
Check Vertex attributes
```{r}
vertex_attr(g4)
```
Check graph attributes (empty names list before adding attributes)
```{r}
graph_attr(g4)
```
Alternative way for adding attributes (set_edge_attr(), set_vertex_attr(), etc. )
```{r}
g4 <- set_graph_attr(g4, "name", "Email Network")
g4 <- set_graph_attr(g4, "something", "A thing")
graph_attr_names(g4)
```
Get individual graph attributes
```{r}
graph_attr(g4, "something")
graph_attr(g4)
```
```{r}
g4 <- delete_graph_attr(g4, "something")
graph_attr(g4)
```
Plot and color vertices by gender
```{r}
#c( "pink", "skyblue")[1+(V(g4)$gender=="male")] # how to select color
plot(g4, edge.arrow.size=.5, vertex.label.color="black", vertex.label.dist=1.5,
     vertex.color=c( "pink", "skyblue")[1+(V(g4)$gender=="male")] )

```

The graph g4 has two edges going from Jim to Jack, and a loop from John to himself. We can simplify our graph to remove loops & multiple edges between the same nodes. Use edge.attr.comb to indicate how edge attributes are to be combined - possible options include sum, mean, prod (product), min, max, first/last (selects the first/last edge’s attribute). Option “ignore” says the attribute should be disregarded and dropped.
```{r}
g4s <- simplify(g4, remove.multiple = T, remove.loops = F, edge.attr.comb=c(weight="sum", type="ignore"))
g4s
plot(g4s, vertex.label.dist=1.5)
```
The description of an igraph object starts with up to four letters:  

  1. D or U, for a directed or undirected graph
  2. N for a named graph (where nodes have a name attribute)
  3. W for a weighted graph (where edges have a weight attribute)
  4. B for a bipartite (two-mode) graph (where nodes have a type attribute)

The two numbers that follow (7 5) refer to the number of nodes and edges in the graph. The description also lists node & edge attributes, for example:  

  * (g/c) - graph-level character attribute
  * (v/c) - vertex-level character attribute
  * (e/n) - edge-level numeric attribute

## Specific graphs and graph models
**Simple graphs**
```{r}
#Empty graph
eg <- make_empty_graph(40)
plot(eg, vertex.size=10, vertex.label=NA)

#full graph
fg <- make_full_graph(40)
plot(fg, vertex.size=10, vertex.label=NA)

# star graph
st <- make_star(40)
plot(st, vertex.size=10, vertex.label=NA)

# Tree graph
tr <- make_tree(60, children =5, mode = "undirected")
plot(tr, vertex.size=10, vertex.label=NA)

# Ring graph
rn <- make_ring(40)
plot(rn, vertex.size=10, vertex.label=NA)

# Erdos-Renyi random graph
er <- sample_gnm(n=100, m=40) # n number of nodes; m number of edges
plot(er, vertex.size=5, vertex.label=NA)
```


**Watts-Strogatz small-world model**

Creates a lattice (with dim dimensions and size nodes across dimension) and rewires edges randomly with probability p. The neighborhood in which edges are connected is nei. You can allow loops and multiple edges.
```{r}
sw <- sample_smallworld(dim=2, size=10, nei=1, p=0.1)
plot(sw, vertex.size=10, vertex.label=NA)
plot(sw, vertex.size=6, vertex.label=NA, layout=layout_in_circle)
plot(sw, vertex.size=6, vertex.label=NA, layout=layout_as_tree)
```
**Barabasi-Albert preferential attachment model for scale-free graphs**  

n is number of nodes, power is the power of attachment (1 is linear); m is the number of edges added on each time step
```{r}
ba <- sample_pa(n=100, power=1, m=1, directed=F)
plot(ba, vertex.size=6, vertex.label=NA)
```

**Notable historical graphs**
```{r}
zach <- graph("Zachary") # Zachary carate club
plot(zach, vertex.size=6, vertex.label=NA)
```
**Rewiring a graph**
each_edge() is a rewiring method that changes the edge endpoints uniformly randomly with a probability prob.
```{r}
plot(rn, vertex.size=6, vertex.label=NA)
rn.rewired <- rewire(rn, each_edge(prob=0.1))
plot(rn.rewired, vertex.size=6, vertex.label=NA)
```
Rewire to connect vertices to other vertices at certain distance
```{r}
rn.neigh <- connect.neighborhood(rn, 5)
plot(rn.neigh, vertex.size=6, vertex.label=NA)
```
Combine graphs (disjoint union, assuming separate vertex sets): `%du%`
```{r}
plot(rn, vertex.size=10, vertex.label=NA)
plot(tr, vertex.size=10, vertex.label=NA)
plot(rn %du% tr, vertex.size=10, vertex.label=NA)

```

# Reading network data
In the following sections of the tutorial, we will work primarily with two small example data sets.
Both contain data about media organizations. One involves a network of hyperlinks and mentions among news sources. The second is a network of links between media venues and consumers. While the example data used here is small, many of the ideas behind the analyses and visualizations we will generate apply to medium and large-scale networks.

## Reading edgelist data
```{r}
nodes <- read.csv("./ruworkshop/Data_files/Dataset1-Media-Example-NODES.csv")
links <- read.csv("./ruworkshop/Data_files/Dataset1-Media-Example-EDGES.csv")
```

Examining data
```{r}
head(nodes)
head(links)

nrow(nodes); length(unique(nodes$id))

#links[,c("from","to")]
nrow(links); nrow(unique(links[,c("from","to")]))
```
Notice that there are more links than unique from-to combinations. That means we have cases in the data where there are multiple links between the same two nodes. We will collapse all links of the same type between the same two nodes by summing their weights, using aggregate() by “from”, “to”, & “type”. We don’t use simplify() here so as not to collapse different link types. E.g. s01 -> s02 appears twice, both time of the type hyperlink. `simplfy()` would collaps all duplicated connections irrespective of their type (possibly without summing their weights)
```{r}
links
#?aggregate # aggregate(x, by, FUN)
# aggregate weight column by from, to, and type columns using the sum function
links <- aggregate(links[,3], links[,-3], sum)
links
links <- links[order(links$from, links$to),] # reorder the table
colnames(links)[4] <- "weight"
links
rownames(links) <- NULL # doesn't appear to be necessary
links

```

## Reading matrix data
Two-mode or bipartite graphs have two different types of actors and links that go across, but not within each type. Our second media example is a network of that kind, examining links between news sources and their consumers.

```{r}
nodes2 <- read.csv("./ruworkshop/Data_files/Dataset2-Media-User-Example-NODES.csv", header = T, as.is = T)
links2 <- read.csv("./ruworkshop/Data_files/Dataset2-Media-User-Example-EDGES.csv", header = T, row.names=1)
```

Examine data
```{r}
head(nodes2)
tail(nodes2)

head(links2)
```

```{r}
links2 <- as.matrix(links2)
links2
dim(links2)
dim(nodes2)
```
## Creating igraph objects
We start by converting the raw data to an igraph network object. Here we use igraph’s `graph.data.frame` function, which takes two data frames: d and vertices.  

  * d describes the edges of the network. Its first two columns are the IDs of the source and the target node for each edge. The following columns are edge attributes (weight, type, label, or anything else).
  * vertices starts with a column of node IDs. Any following columns are interpreted as node attributes.

### Dataset 1
```{r}
net <- graph_from_data_frame(d=links, vertices=nodes, directed=T)
class(net)
net
```

```{r}
plot(net, edge.arrow.size=.4, vertex.label=NA)
```
Cleaning up the network
```{r}
net <- simplify(net, remove.multiple = F, remove.loops=T)
plot(net, edge.arrow.size=.4,vertex.label=NA)
```
You might notice that we could have used simplify to combine multiple edges by summing their weights with a command like `simplify(net, edge.attr.comb=list(weight="sum","ignore"))`. The problem is that this would also combine multiple edge types (in our data: “hyperlinks” and “mentions”).  

If you need them, you can extract an edge list or a matrix from igraph networks. In graph theory and computer science, an adjacency matrix (=connection matrix) is a square matrix used to represent a finite graph. The elements of the matrix indicate whether pairs of vertices are adjacent or not in the graph.
```{r}
as_edgelist(net, names=T)
as_adjacency_matrix(net, attr="weight")
#?as_adjacency_matrix
```

Or data frames describing nodes and edges:
```{r}
as_data_frame(net, what="edges")
as_data_frame(net, what="vertices")
```
### Dataset 2
As we have seen above, this time the edges of the network are in a matrix format. We can read those into a graph object using graph_from_incidence_matrix(). In igraph, bipartite networks have a node attribute called type that is FALSE (or 0) for vertices in one mode and TRUE (or 1) for those in the other mode.
```{r}
head(nodes2)
tail(nodes2)

#head(links2)
links2
```
```{r}
net2 <- graph_from_incidence_matrix(links2) # for bipartite networks
table(V(net2)$type) #here, nodes of type s are F (or 0) and nodes of type U are T (or 1)

```
To transform a one-mode network matrix into an igraph object, use instead `graph_from_adjacency_matrix()`.  
We can also easily generate bipartite projections for the two-mode network: (co-memberships are easy to calculate by multiplying the network matrix by its transposed matrix, or using igraph’s bipartite.projection() function).  
[Bipartite networks](https://en.wikipedia.org/wiki/Bipartite_network_projection) are a particular class of complex networks, whose nodes are divided into two sets X and Y, and only connections between two nodes in different sets are allowed. For the convenience of directly showing the relation structure among a particular set of nodes, bipartite networks are usually compressed by one-mode projection. This means that the ensuing network contains nodes of only either of the two sets, and two X (or, alternatively, Y) nodes are connected only if when they have at least one common neighboring Y (or, alternatively, X) node.
```{r}
net2.bp <- bipartite.projection(net2)
net2.bp
```

Projections can be calculated manually as well.
```{r}
as_incidence_matrix(net2)
t(as_incidence_matrix(net2))

as_incidence_matrix(net2) %*% t(as_incidence_matrix(net2)) # projection 1
t(as_incidence_matrix(net2)) %*% as_incidence_matrix(net2) # projection 2
```
**Projection 1**
```{r}
plot(net2.bp$proj1, vertex.label.color="black", vertex.label.dist=1,
     vertex.size=7, vertex.label=nodes2$media[!is.na(nodes2$media.type)])
```
**Projection 2**
```{r}
plot(net2.bp$proj2, vertex.label.color="black", vertex.label.dist=1, 
     vertex.size=7, vertex.label=nodes2$media[ is.na(nodes2$media.type)])
```

# Plotting network data

Plotting with igraph: the network plots have a wide set of parameters you can set. Those include node options (starting with vertex.) and edge options (starting with edge.). A list of selected options is included below, but you can also check out `?igraph.plotting` for more information.  

## Plotting parameters

|**NODES**           |                 |
|:-------------------|:----------------|
|vertex.color        |     Node color  |
|vertex.frame.color  | Node border color |
|vertex.shape        | One of “none”, “circle”, “square”, “csquare”, “rectangle”, “crectangle”, “vrectangle”, “pie”, “raster”, or “sphere”|
|vertex.size         | Size of the node (default is 15)|
|vertex.size2        | The second size of the node (e.g. for a rectangle)|
|vertex.label        | Character vector used to label the nodes|
|vertex.label.family | Font family of the label (e.g.“Times”, “Helvetica”)|
|vertex.label.font   | Font: 1 plain, 2 bold, 3, italic, 4 bold italic, 5 symbol|
|vertex.label.cex    | Font size (multiplication factor, device-dependent)|
|vertex.label.dist   | Distance between the label and the vertex|
|vertex.label.degree | The position of the label in relation to the vertex, where 0 right, “pi” is left, “pi/2” is below, and “-pi/2” is above|
|**EDGES**         |                    |
|edge.color        | Edge color |
|edge.width        | Edge width, defaults to 1 |
|edge.arrow.size   | Arrow size, defaults to 1 |
|edge.arrow.width  | Arrow width, defaults to 1 |
|edge.lty          | Line type, could be 0 or “blank”, 1 or “solid”, 2 or “dashed”, 3 or “dotted”, 4 or “dotdash”, 5 or “longdash”, 6 or “twodash” |
|edge.label        | Character vector used to label edges |
|edge.label.family | Font family of the label (e.g.“Times”, “Helvetica”) |
|edge.label.font   | Font: 1 plain, 2 bold, 3, italic, 4 bold italic, 5 symbol |
|edge.label.cex    | Font size for edge labels |
|edge.curved       | Edge curvature, range 0-1 (FALSE sets it to 0, TRUE to 0.5) |
|arrow.mode        | Vector specifying whether edges should have arrows, possible values: 0 no arrow, 1 back, 2 forward, 3 both |
|**OTHER** |  |
|margin | Empty space margins around the plot, vector with length 4 |
|frame  | if TRUE, the plot will be framed |
|main   | If set, adds a title to the plot |
|sub    | If set, adds a subtitle to the plot |

We can set the node & edge options in two ways - the first one is to specify them in the plot()
function, as we are doing below.
```{r}
plot(net, edge.arrow.size=.4, edge.curved=.1)
```
```{r}
# Set edge color to gray, and the node color to orange.
# Replace the vertex label with the node names stored in "media"

# as_data_frame(net, what = "both")
# V(net)$media
# V(net)$type.label

plot(net, edge.arrow.size=.2, edge.curved=0, 
     vertex.color="orange", vertex.frame.color="#555555",
     vertex.label=V(net)$media, vertex.label.color="black",
     vertex.label.cex=.7)

#plot(net, edge.arrow.size=.2, edge.curved=0, vertex.color="orange", vertex.frame.color="#555555", vertex.label=V(net)$type.label, vertex.label.color="black", vertex.label.cex=.7)

```
The second way to set attributes is to add them to the igraph object. Let’s say we want to color our network nodes based on type of media, and size them based on audience size (larger audience -> larger node). We will also change the width of the edges based on their weight.
```{r}
# Generate colors based on media type:
colrs <- c("gray50", "tomato", "gold")
```

The attributes added to the network need to have specific names and don't have to be specified again in the `plot()` function.
```{r}
# Select colors by media type
#V(net)$media.type # access vertex media type attribute. 
V(net)$color <- colrs[V(net)$media.type]

# Set node size based on audience size:
V(net)$size <- V(net)$audience.size*0.7
vertex_attr(net)

# The labels are currently node IDs
# Setting them to NA will render no labels:
V(net)$label.color <- "black"
V(net)$label <- NA

# change arrow size and edge color:
E(net)$arrow.size <- .2
E(net)$edge.color <- "gray80"

E(net)$width <- 1+E(net)$weight/12
#edge_attr(net)
plot(net)
```
The network layout can be adjusted as well
```{r}
graph_attr(net, "layout") <- layout_with_lgl
plot(net)
```
Attributes from the network can be overwritten in the `plot()` function
```{r}
plot(net, edge.color="orange", vertex.color="gray50")
```
Adding a legend
```{r}
plot(net)
legend(x=-1.5, y=-1.1, c("Newspaper", "Television", "Online News"), pch=21, col="gray90", pt.bg = colrs, pt.cex=2, cex = .8, ncol=1)
```
For semantic networks, we may be interested in plotting only the labels of the nodes:
```{r}
plot(net, vertex.shape="none", vertex.label=V(net)$media, vertex.label.font=2, vertex.label.color="gray40", vertex.label.cex=.7, edge.color="gray85")
```
Let’s color the edges of the graph based on their source node color. We can get the starting node for each edge with the `ends()` igraph function.
```{r dpi=500}
#ends(net, es=E(net), names=T)
edge.start <- ends(net, es=E(net), names=F)[,1]
edge.col <- V(net)$color[edge.start]
plot(net, edge.color=edge.col, edge.curved=.2)
```

## Network Layouts
Network layouts are simply algorithms that return coordinates for each node in a network. For the purposes of exploring layouts, we will generate a slightly larger 80-node graph. We use the sample_pa() function which generates a simple graph starting from one node and adding more nodes and links based on a preset level of preferential attachment (Barabasi-Albert model).
```{r}
net.bg <- sample_pa(80)
V(net.bg)$size <- 8
V(net.bg)$frame.color <- "white"
V(net.bg)$color <- "orange"
V(net.bg)$label <- ""
E(net.bg)$arrow.mode <- 0

plot(net.bg)
```
Setting the layout in the plot function
```{r}
plot(net.bg,layout=layout_randomly)
```
Calculating the vertex coordinates in advance
```{r}
#l <- layout_randomly(net.bg) # random layout
l <- layout_in_circle(net.bg)
plot(net.bg, layout=l)

```
l1 and l2 are matrices of x, y coordinates (N x 2) for the N nodes in the graph. You can easily
generate your own:
```{r}
#vcount(net.bg) # count vertices
l <- cbind(1:vcount(net.bg), c(1,vcount(net.bg):2)) # generate x,y coordinates for 80 vertices
plot(net.bg, layout=l)
```
Further build-in layouts
```{r}
par(mfrow=c(1,3), mar=c(0,0,0,0))
# Randomly placed vertices
l <- layout_randomly(net.bg)
plot(net.bg, layout=l)

# Circle layout
l <- layout_in_circle(net.bg)
plot(net.bg, layout=l)

# 3D sphere layout
l <- layout_on_sphere(net.bg)
plot(net.bg, layout=l)

```
Fruchterman-Reingold is one of the most used force-directed layout algorithms out there.  

Force-directed layouts try to get a nice-looking graph where edges are similar in length and cross each other as little as possible. They simulate the graph as a physical system. Nodes are electrically charged particles that repulse each other when they get too close. The edges act as springs that attract connected nodes closer together. As a result, nodes are evenly distributed through the chart area, and the layout is intuitive in that nodes which share more connections are closer to each other. The disadvantage of these algorithms is that they are rather slow and therefore less often used in graphs larger than ~1000 vertices. You can set the “weight” parameter which increases the attraction forces among nodes connected by heavier edges.
```{r}
l <- layout_with_fr(net.bg)
plot(net.bg, layout=l)
```
You will notice that the layout is not deterministic - different runs will result in slightly different configurations. Saving the layout in l allows us to get the exact same result multiple times, which can be helpful if you want to plot the time evolution of a graph, or different relationships – and want nodes to stay in the same place in multiple plots.
```{r}
par(mfrow=c(2,2), mar=c(0,0,0,0)) # plot four figures - 2 rows, 2 columns
plot(net.bg, layout=layout_with_fr)
plot(net.bg, layout=layout_with_fr)
plot(net.bg, layout=l)
plot(net.bg, layout=l)
```
By default, the coordinates of the plots are rescaled to the [-1,1] interval for both x and y. You can change that with the parameter rescale=FALSE and rescale your plot manually by multiplying the coordinates by a scalar. You can use norm_coords to normalize the plot with the boundaries you want.
```{r}
l <- layout_with_fr(net.bg)
l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)

par(mfrow=c(2,2), mar=c(0,0,0,0))
plot(net.bg, rescale=F, layout=l*0.4)
plot(net.bg, rescale=F, layout=l*0.6)
plot(net.bg, rescale=F, layout=l*0.8)
plot(net.bg, rescale=F, layout=l*1.0)
```

Another popular force-directed algorithm that produces nice results for connected graphs is Kamada Kawai. Like Fruchterman Reingold, it attempts to minimize the energy in a spring system.
```{r}
l <- layout_with_kk(net.bg)
plot(net.bg, layout=l)
```
All available layouts in igraph:
```{r}
layouts <- grep("^layout_", ls("package:igraph"), value=TRUE)[-1]
# Remove layouts that do not apply to our graph.
layouts <- layouts[!grepl("bipartite|merge|norm|sugiyama|tree", layouts)]
layouts
```
```{r}
par(mfrow=c(3,3), mar=c(1,1,1,1))
for (layout in layouts) {
#print(layout)
l <- do.call(layout, list(net))
plot(net, edge.arrow.mode=0, layout=l, main=layout) }
```

## Interactive plotting with tkplot
R and igraph allow for interactive plotting of networks. This might be a useful option for you if you want to tweak slightly the layout of a small graph. After adjusting the layout manually, you can get the coordinates of the nodes and use them for other plots.
```{r}
tkid <- tkplot(net) #tkid is the id of the tkplot that will open
l <- tkplot.getcoords(tkid) # grab the coordinates from tkplot
tk_close(tkid, window.close = T)
plot(net, layout=l)
```
# Network and node descriptives
## Density

The proportion of present edges from all possible edges in the network.
```{r}
#plot(net)
edge_density(net, loops=F)
ecount(net)/(vcount(net)*(vcount(net)-1)) #for a directed network
```
## Reciprocity
The proportion of reciprocated ties (for a directed network).
```{r}
reciprocity(net)
dyad_census(net) # Mutual, asymmetric, and null node pairs
2*dyad_census(net)$mut/ecount(net) # Calculating reciprocity
```


## Transitivity
  * global - ratio of triangles (direction disregarded) to connected triples.
  * local - ratio of triangles to connected triples each vertex is part of.
```{r}
transitivity(net, type="global") # net is treated as an undirected network
transitivity(as.undirected(net, mode="collapse")) # same as above
```

```{r}
transitivity(net, type="local")
triad_census(net) # for directed networks
```

A network diameter is the longest geodesic distance (length of the shortest path between two nodes) in the network. In igraph, diameter() returns the distance, while get_diameter() returns the nodes along the first found path of that distance.  
Note that edge weights are used by default, unless set to NA.
```{r}
diameter(net, directed=F, weights=NA)
```

```{r}
diameter(net, directed=F)
```
```{r}
diam <-  get_diameter(net, directed=T)
diam
```
Note that get_diameter() returns a vertex sequence. Note though that when asked to behaved as a vector, a vertex sequence will produce the numeric indexes of the nodes in it. The same applies for edge sequences.

```{r}
class(diam)
```
```{r}
as.vector(diam)
```
Color nodes along the diameter:
```{r}
vcol <- rep("gray40", vcount(net))
vcol[diam] <- "gold"
ecol <- rep("gray80", ecount(net))
ecol[E(net, path=diam)] <- "orange"
# E(net, path=diam) finds edges along a path, here 'diam'
plot(net, vertex.color=vcol, edge.color=ecol, edge.arrow.mode=0)
```
## Node degrees
The function degree() has a mode of in for in-degree, out for out-degree, and all or total for total degree.
```{r}
deg <- degree(net, mode="all")
plot(net, vertex.size=deg*4) # nodes/vertices with higher degrees are larger
```
## Dgree distribution

```{r}
deg.dist <- degree_distribution(net, cumulative=T, mode="all")
plot( x=0:max(deg), y=1-deg.dist, pch=19, cex=1.2, col="orange", xlab="Degree", ylab="Cumulative Frequency")
```

## Centrality & centralization
Centrality functions (vertex level) and centralization functions (graph level). The centralization functions return `res` - vertex centrality, `centralization`, and `theoretical_max` - maximum centralization score for a graph of that size. The centrality function can run on a subset of nodes (set with the `vids` parameter). This is helpful for large graphs where calculating all centralities may be a resource-intensive and time-consuming task.  


*Degree* (number of ties)
```{r}
degree(net, mode="in")
centr_degree(net, mode="in", normalized=T)
```
*Closeness* (centrality based on distance to others in the graph)  
Inverse of the node's average geodesic distance to others in the network
```{r}
closeness(net, mode="all", weights=NA)
centr_clo(net, mode="all", normalized=T)
```

*Eigenvector* (centrality proportional to the sum of connection centralities)
Values of the first eigenvector of the graph matrix.
```{r}
eigen_centrality(net, directed=T, weights=NA)
centr_eigen(net, directed=T, normalized=T)
```
Betweenness (centrality based on a broker position connecting others)  
Number of geodesics that pass through the node or the edge.
```{r}
betweenness(net, directed=T, weights=NA)
edge_betweenness(net, directed=T, weights=NA)
centr_betw(net, directed=T, normalized=T)
```

# Exploring the network data
## Hubs and authorities
The hubs and authorities algorithm developed by Jon Kleinberg was initially used to examine web pages. Hubs were expected to contain catalogs with a large number of outgoing links; while authorities would get many incoming links from hubs, presumably because of their high-quality relevant information.
```{r}
hs <- hub_score(net, weights=NA)$vector
as <- authority_score(net, weights=NA)$vector
l <- layout_with_gem(net)

par(mfrow=c(1,2))
plot(net, vertex.size=hs*50, layout=l, main="Hubs")
plot(net, vertex.size=as*50, layout=l, main="Authorities")
```
For undirected matrices the adjacency matrix is symmetric and the hub scores are the same as authority scores, see [authority_score](https://igraph.org/r/doc/authority_score.html) or [hub_score](https://igraph.org/r/doc/hub_score.html).


## Distances and paths
Average path length: the mean of the shortest distance between each pair of nodes in the network (in both directions for directed graphs).
```{r}
mean_distance(net, directed=F)
```


```{r}
mean_distance(net, directed=T)
```
We can also find the length of all shortest paths in the graph:
```{r}
distances(net) # with edge weights
distances(net, weights=NA) # ignore weights
```
We can extract the distances to a node or set of nodes we are interested in. Here we will get the distance of every media from the New York Times.
```{r}
dist.from.NYT <- distances(net, v=V(net)[media=="NY Times"], to=V(net), weights=NA)
# Set colors to plot the distances:
oranges <- colorRampPalette(c("dark red", "gold"))
col <- oranges(max(dist.from.NYT)+1)
col <- col[dist.from.NYT+1]
plot(net, vertex.color=col,vertex.label=dist.from.NYT, edge.arrow.size=.6, vertex.label.color="white")
```
We can also find the shortest path between specific nodes. Say here between MSNBC and the New York Post:
```{r}
news.path <- shortest_paths(net, from = V(net)[media=="MSNBC"], 
                            to = V(net)[media=="New York Post"], 
                            output = "both") # both path nodes and edges

# Generate edge color variable to plot the path:
ecol <- rep("gray80", ecount(net))
ecol[unlist(news.path$epath)] <- "orange"

# Generate edge width variable to plot the path:
ew <- rep(2, ecount(net))
ew[unlist(news.path$epath)] <- 4

# Generate node color variable to plot the path:
vcol <- rep("gray40", vcount(net))
vcol[unlist(news.path$vpath)] <- "gold" 
plot(net, vertex.color=vcol, edge.color=ecol, edge.width=ew, edge.arrow.mode=0)
```
Identify the edges going into or out of a vertex, for instance the WSJ. For a single node, use `incident()`, for multiple nodes use `incident_edges()`
```{r}
inc.edges <- incident(net, V(net)[media=="Wall Street Journal"], mode="all")

# Set colors to plot the selected edges.
ecol <- rep("gray80", ecount(net))
ecol[inc.edges] <- "orange"

vcol <- rep("grey40", vcount(net))
vcol[V(net)$media=="Wall Street Journal"] <- "gold"

plot(net, vertex.color=vcol, edge.color=ecol)
```
We can also easily identify the immediate neighbors of a vertex, say WSJ. The `neighbors` function finds all nodes one step out from the focal actor. To find the neighbors for multiple nodes, use `adjacent_vertices()` instead of `neighbors()`. To find node neighborhoods going more than one step out, use function `ego()` with parameter `order` set to the number of steps out to go from the focal node(s).
```{r}
neigh.nodes <- neighbors(net, V(net)[media=="Wall Street Journal"], mode="out")

# Set colors to plot the neighbors:
vcol[neigh.nodes] <- "#ff9d00"
plot(net, vertex.color=vcol)
```
Special operators for the indexing of edge sequences: %–%, %->%, %<-%  

  * E(network)[X %–% Y] selects edges between vertex sets X and Y, ignoring direction
  * E(network)[X %->% Y] selects edges from vertex sets X to vertex set Y
  * E(network)[X %->% Y] selects edges from vertex sets Y to vertex set X

For example, select edges from newspapers to online sources:
```{r}
E(net)[ V(net)[type.label=="Newspaper"] %->% V(net)[type.label=="Online"] ]
```
Co-citation (for a couple of nodes, how many shared nominations they have):
```{r}
cocitation(net)
```
## Subgroups and communities
Before we start, we will make our network undirected. There are several ways to do that conversion:

  * We can create an undirected link between any pair of connected nodes (`mode="collapse"`)
  * Create undirected link for each directed one in the network, potentially ending up with a multiplex graph (`mode="each"`)
  * Create undirected link for each symmetric link in the graph (`mode="mutual"`).

In cases when we may have ties *A -> B* and *B -> A* ties collapsed into a single undirected link, we need to specify what to do with their edge attributes using the parameter `edge.attr.comb` as we did earlier with `simplify()`. Here we have said that the ‘weight’ of the links should be summed, and all other edge attributes ignored and dropped.
```{r}
net.sym <- as.undirected(net, mode= "collapse", edge.attr.comb=list(weight="sum", "ignore"))
```

### Cliques
Find cliqus (complete subgraphs of an undirected graph)
```{r}
#cliques(net.sym) # list of cliques
sapply(cliques(net.sym), length)
largest_cliques(net.sym)
```
```{r}
vcol <- rep("grey80", vcount(net.sym))
vcol[unlist(largest_cliques(net.sym))] <- "gold"
plot(as.undirected(net.sym), vertex.label=V(net.sym)$name, vertex.color=vcol)
```
### Community detection
A number of algorithms aim to detect groups that consist of densely connected nodes with fewer connections across groups.  

#### Community detection based on edge betweenness (Newman-Girvan)**  
High-betweenness edges are removed sequentially (recalculating at each step) and the best partitioning of the network is selected.
```{r}
ceb <- cluster_edge_betweenness(net)
#ceb
```

```{r}
dendPlot(ceb, mode="hclust")
```


```{r}
plot(ceb, net)
```
Let’s examine the community detection igraph object:
```{r}
class(ceb)
```

```{r}
length(ceb) # number of communities
```
```{r}
membership(ceb) # community membership for each node
```
```{r}
modularity(ceb) # how modular the graph partitioning is
```
```{r}
crossing(ceb, net) # boolean vector: TRUE for edges across communities
```
High modularity for a partitioning reflects dense connections within communities and sparse
connections across communities.

#### Community detection based on based on propagating labels  

Assigns node labels, randomizes, than replaces each vertex’s label with the label that appears most frequently among neighbors. Those steps are repeated until each vertex has the most common label of its neighbors.
```{r message=FALSE, warning=FALSE, echo=FALSE}
clp <- cluster_label_prop(net)
plot(clp, net)
```
#### Community detection based on greedy optimization of modularity
```{r}
cfg <- cluster_fast_greedy(as.undirected(net))
plot(cfg, as.undirected(net))
```
We can also plot the communities without relying on their built-in plot:
```{r}
V(net)$community <- cfg$membership
colrs <- adjustcolor( c("gray50", "tomato", "gold", "yellowgreen"), alpha=.6)
plot(net, vertex.color=colrs[V(net)$community])
```

**Note**  
In order to highlight certain nodes is possible to draw a polygon around them using the `mark.groups` argument of the `plot.igraph` function. Check this [blog](https://assemblingnetwork.wordpress.com/2013/05/29/network-basics-with-r-and-igraph-part-i-of-iii/) for examples.


### K-core decomposition
The k-core is the maximal subgraph in which every node has degree of at least k. This also means that the (k+1)-core will be a subgraph of the k-core. The result here gives the coreness of each vertex in the network. A node has coreness D if it belongs to a D-core but not to (D+1)-core.
```{r}
kc <- coreness(net, mode="all")
plot(net, vertex.size=kc*6, vertex.label=kc, vertex.color=colrs[kc])
```
## Assortativity and homophily

Homophily: the tendency of nodes to connect to others who are similar on some variable.

  * assortativity_nominal() is for categorical variables (labels)
  * assortativity() is for ordinal and above variables
  * assortativity_degree() checks assortativity in node degrees

```{r}
assortativity_nominal(net, V(net)$media.type, directed=F)
```
 
Matching of attributes across connected nodes more than expected by chance
```{r}
assortativity(net, V(net)$audience.size, directed=F)
```

Correlation of attributes across connected nodes
```{r}
assortativity_degree(net, directed=F)
```
As above, with the focal attribute being the node degree D-1


